Cultural Transmission Can Explain the Evolution of Cooperative Behavior

### 2020-07-13 ms/main.tex
reorganizes results 3-4

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
If 
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)]
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of the cooperator phenotype $A$ among parents increases every generation.

@@

\textbf{Result 2: Oblique and horizontal transmission of cooperation.} The frequency of phenotype $A$ among adults increases if $p'>p$.
Using eq.~\ref{eq:nextgen_parents_oblique_only}, we find that $p'>p$ occurs when
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B . 
\end{equation}
Therefore, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~\textbf{\autoref{fig:results_c}}).

@@

In this case $0<v<1$, and the equation system is more complex than before.
Therefore, we focus on local rather than global stability, which was the case so far.\\
To proceed, we note that 
eq.~\ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
eq.~\ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\hat{p}$, and 
eq.~\ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$.
We then determine the equilibria of this equation, that is, solutions for $\hat{p}' = \hat{p}$, and their stability.

\textbf{Result 3: Oblique and vertical transmission with symmetric horizontal transmission.}

For simplicity, we start with the assumption that $T_A = T_B = T$. We have
\begin{equation} \label{eq:equal_horizontal_transmission}
% DOR: I checked again, substituting eq 7 and 9 to eq 5: you have to divide the RHS by wbar.
  %\hat{p}' - \hat{p} = 
  %\hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T)) = 
  \bar{w}(\hat{p}' - \hat{p}) = 
  f(\hat{p}) = \hat{p}(1-\hat{p})(\alpha bvT - cv(1-T)) % shorter version
\end{equation}

We are looking for the equilibrium $\hat{p}^*$ which occurs when $f(\hat{p}^*)=0$, or $\hat{p}' = \hat{p}$.
It is easy to verify that fixation of either phenotype is an equilibrium, that is,
\begin{equation} \label{eq:result2.1}
  \hat{p}^*_0 =  0, \, \hat{p}^*_1 = 1 .
\end{equation}

Is fixation of cooperation, $\hat{p}^*_1=1$, locally stable?. 
An equilibrium point $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})$ at $p^*$ is negative, $f'(\hat{p}^*)<0$. 
This derivative is
\begin{equation} \label{eq:derivative_of_phattag-phat}
	f'(\hat{p}) \mid_{\hat{p}=1} = 
% 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
	(1- 2\hat{p}) (\alpha bvT - cv(1-T)) \mid_{\hat{p}=1} = 
	-\alpha bvT + cv(1-T)).
\end{equation}
Therefore, fixation of cooperation is locally stable when
\begin{equation} \label{eq:oblique_and_vertic_result2}
  %\frac{b}{c}>\frac{1-T}{\alpha T}
	c < b \cdot \frac{\alpha T}{1-T},
\end{equation}
which is the same condition we found in Corollary 1.1, Eq.~\ref{eq:equal_transmission}.

\textbf{Result 4: Oblique and vertical transmission of cooperation with asymmetric horizontal transmission.}

We now study the general case where $T_A \neq T_B$. We have
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) \\
%\beta_2 &= -[c(1-v) - b(1-\alpha v) +  1](T_A-T_B)-\alpha bv T_A + cv(1-T_B) \\
\beta_2 &= -\beta_1 -\beta_3  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}
Since $f(\hat{p})$ is a cubic polynomial, three equilibria exist.
We already know that $\hat{p}^*_0 = 0 $ and $\hat{p}^*_1 = 1$ are equilibria.
The third equilibrium $\hat{p}^*_2$ can be found by solving $f(\hat{p})/\big[\hat{p}(1-\hat{p})\big] = \beta_3 -\beta_1 \hat{p} = 0$:
\begin{equation} \label{eq:oblique_and_vertic_result}
  \hat{p}^*_2 =  
  \frac{\beta_3}{\beta_1}.
  %\frac{\alpha bvT_A - cv(1-T_B) + (T_A-T_B)}
  %     {\big(c(1-v) -b (1-\alpha v)\big) (T_A-T_B)} .
\end{equation}
To determine when $\hat{p}^*_2$ is a legitimate equilibrium (i.e. $0<\hat{p}^*_2<1$), we first focus on the left inequality, $0<\hat{p}^*_2$, and
%To do that we find the derivative  of $f(\hat{p})$,
%\begin{equation} \label{eq:derivative_of_phattag-phat2}
%  f'(\hat{p}) = 3\beta_1 \hat{p}^2 + 2\beta_2\hat{p}^2 + \beta_3
% \end{equation}
assume that $T_A>T_B$. In this case,
\begin{equation} \label{eq:beta1}
  \begin{split}
   \beta_1 & = [c(1-v) - b (1-\alpha v)] (T_A-T_B) \\
   &< [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) \\
   &= (1-\alpha v)(c-b)(T_A-T_B) < 0 ,
  \end{split}
 \end{equation}
since $T_A > T_B$, $c<b$, and $1>\alpha v$.
Therefore, when $\hat{p}$ goes to negative infinity, the cubic $f(\hat{p})$ goes to positive infinity.
So, a necessary condition for $\hat{p}^*_2>0$ is that $f'(0) = \beta_3 < 0$.
This condition also satisfies local stability of $\hat{p}^*_0=0$.
Hence, if $\hat{p}^*_2$ is legitimate, then extinction of cooperation is locally stable.

We now focus on $\hat{p}^*_2<1$, still assuming $T_A>T_B$.
When $\hat{p}$ goes to positive infinity, the cubic $f(\hat{p})$ goes to negative infinity (Eq.~\ref{eq:beta1}).
So, a necessary condition for $\hat{p}^*_2<1$ is that $f'(1)=\beta_1 - \beta_3<0$.
This condition also satisfies local stability of $\hat{p}^*_1=1$.
Hence, if $\hat{p}^*_2$ is legitimate, then fixation of cooperation is also locally stable.
Moreover, since legitimacy of $\hat{p}^*_2$ implies local stability of both $\hat{p}^*_0$ and $\hat{p}^*_1$, we conclude that $\hat{p}^*_2$ is locally unstable.

If we assume instead that $T_A < T_B$, we reach the following results:

\textbf{Result 4}: if $T_A>T_B$ ($T_B>T_A$), then $\hat{p}^*_2$ is a legitimate locally unstable equilibrium when $\beta_1<\beta_3<0$ ($\beta_1>\beta_3>0$). In this case, both fixations are locally stable and $\hat{p}^*_2$ separates the domains of attraction.

\begin{figure}[H]
  \centering
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2a.pdf}
    \caption{$v=1$, $T_A=T_B=T$, $\alpha \neq 0$}
    \label{fig:results_a}
  \end{subfigure}
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2b.pdf}
    \caption{$v=1$, $\alpha = 0$}
    \label{fig:results_b}
  \end{subfigure}
  \begin{subfigure}{8cm}
    \includegraphics[scale=0.5]{figure2c.pdf}
    \caption{$v=0$}
    \label{fig:results_c}
  \end{subfigure}
  \label{fig:results}
  \caption{
  \textbf{Numerical results for cultural evolution of cooperation.}
  Shown are dynamics of \textbf{(a-b)} $\tilde{p}$, the frequency of parents with cooperative phenotype $A$; \textbf{(c)} $p'$, the frequency of adults with cooperative phenotype $A$.
  The figure demonstrates fixation of cooperation (green), extinction of cooperation (blue)m and stable co-existence of cooperators and defectors (orange). %TODO ic (c), the blue and green got mixed. 
  % TODO have all panels in one row.
  }
\end{figure}

>>>>

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
If 
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)] ,
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of the cooperator phenotype $A$ among parents increases every generation.

@@

\textbf{Result 2: Oblique and horizontal transmission of cooperation.} 
If there is a horizontal transmission bias in favor of cooperation,
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B, 
\end{equation}
then $p'>p$, and the frequency of the cooperator phenotype among adults increases every generation.
Therefore, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~\textbf{\autoref{fig:results_c}}).

@@

In this case $0<v<1$, and the equation system is more complex.
Therefore, we focus on local, rather than global, stability.\\
To proceed, we note that 
eq.~\ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$,
eq.~\ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\hat{p}$, and 
eq.~\ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$.
We then determine the equilibria of this equation, that is, solutions for $\hat{p}' = \hat{p}$, and their stability.

We start with the simpler case of symmetric horizontal transmission, $T=T_A=T_B$. 
We apply eqs.~\ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} to define $f(\hat{p})$ as 
% Yoav: I checked again, substituting eq 7 and 9 to eq 5: you have to divide the RHS by wbar.
\begin{equation} \label{eq:equal_horizontal_transmission}
\begin{aligned}
  %\hat{p}' - \hat{p} = 
  %\hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T)) = 
  f(\hat{p}) &= 
  \bar{w}(\hat{p}' - \hat{p}) = \\
  &\hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big]. % shorter version
\end{aligned}
\end{equation}

The equilibria $\hat{p}^*$ are solutions to $f(\hat{p}^*)=0$, or $\hat{p}'^* = \hat{p}^*$.
It is easy to verify that fixation of either phenotype is an equilibrium, that is,
\begin{equation} \label{eq:result2.1}
  \hat{p}^*_0 =  0, \, \hat{p}^*_1 = 1 .
\end{equation}
An equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})$ at $p^*$ is negative, $f'(\hat{p}^*)<0$. 
Since the derivative is 
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
then the condition for local stability of $\hat{p}^*_1$ is
\begin{equation} \label{eq:derivative_of_phattag-phat}
	f'(1) =
% 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
%	(1- 2\hat{p}) (\alpha bvT - cv(1-T)) \mid_{\hat{p}=1} = 
	-\alpha bvT + cv(1-T) < 0,
\end{equation}
which gives us the following result.

\textbf{Result 3: Oblique and vertical transmission with symmetric horizontal transmission.}
If horizontal transmission is symmetric, $T = T_A = T_B$, and if
\begin{equation} \label{eq:oblique_and_vertic_result2}
  %\frac{b}{c}>\frac{1-T}{\alpha T}
	c < b \cdot \frac{\alpha T}{1-T},
\end{equation}
then fixation of the cooperator phenotype $A$ is locally stable.
This is the same condition we found in Corollary 1.1, Eq.~\ref{eq:equal_transmission}.
\\
\\
We now turn to the general case where $T_A \neq T_B$. We have
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) \\
%\beta_2 &= -[c(1-v) - b(1-\alpha v) +  1](T_A-T_B)-\alpha bv T_A + cv(1-T_B) \\
\beta_2 &= -\beta_1 -\beta_3  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}
Since $f(\hat{p})$ is a cubic polynomial, three equilibria exist.
We already know that $\hat{p}^*_0 = 0 $ and $\hat{p}^*_1 = 1$ are equilibria.
By solving $f(\hat{p})/\big[\hat{p}(1-\hat{p})\big] = \beta_3 -\beta_1 \hat{p} = 0$ we can find the third equilibrium,
\begin{equation} \label{eq:oblique_and_vertic_result}
  \hat{p}^*_2 =  
  \frac{\beta_3}{\beta_1}.
  %\frac{\alpha bvT_A - cv(1-T_B) + (T_A-T_B)}
  %     {\big(c(1-v) -b (1-\alpha v)\big) (T_A-T_B)} .
\end{equation}

Note that the sign of the cubic $f(\hat{p})$ at positive (negative) infinity is equal (inverse) to the sign of $\beta_1$. 
If $T_A>T_B$, then 
\begin{equation} \label{eq:beta1}
   \beta_1 %& = [c(1-v) - b (1-\alpha v)] (T_A-T_B) \\
   < [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) 
   = (1-\alpha v)(c-b)(T_A-T_B) < 0 ,
 \end{equation}
since $c<b$, and $1>\alpha v$.
So the sign of the cubic at positive and negative infinity is negative and positive, respectively.
First, if $\beta_3<\beta_1$ then $0<\hat{p}^*_2<1$ and therefore $f'(0)<0$ and $f'(1)<0$, that is, both fixations are locally stable and $\hat{p}^*_2$ separates the domains of attraction.
Second, if $\beta_1<\beta_3<0$ then $1<\hat{p}^*_2$ and therefore $f'(0)<0$ and $f'(1)>0$, that is, fixation of the defector phenotype $B$ is the only locally stable legitimate (i.e. between 0 and 1) equilibrium.
Third, if $0<\beta_3$ then $\hat{p}^*_2<0$ and therefore $f'(0)>0$ and $f'(1)<0$, that is, fixation of the cooperator phenotype $A$ is the only locally stable legitimate equilibrium.

Similarly, if $T_B>T_A$, then
\begin{equation} \label{eq:beta1_rev}
   \beta_1 %& = -[c(1-v) - b (1-\alpha v)] (T_A-T_B) \\
   > [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) 
   = (1-\alpha v)(c-b)(T_A-T_B) > 0,
 \end{equation}
since $c<b$, and $1>\alpha v$. So the sign of the cubic at positive and negative infinity is positive and negative, respectively. 
First, if $\beta_3<0$ then $\hat{p}^*_2<0$ and therefore $f'(0)<0$ and $f'(1)>0$, that is, fixation of the defector phenotype $A=B$ is the only locally stable legitimate equilibrium.
Second, if $0<\beta_3<\beta_1$ then $0<\hat{p}^*_2<1$ and therefore $f'(0)>0$ and $f'(1)>0$, that is, both fixations are locally unstable and $\hat{p}^*_2$ is a stable polymorphic equilibrium.
Third, if $\beta_1<\beta_3$ then $\hat{p}^*_2>1$ and therefore $f'(0)>0$ and $f'(1)<0$, that is, fixation of the cooperator phenotype $A$ is the only locally stable legitimate equilibrium.

The following result summarizes these findings.

\textbf{Result 4: Oblique and vertical transmission of cooperation with asymmetric horizontal transmission.}
The cultural evolution of a cooperator phenotype will follow on of th following scenarios, depending on the signs of the horizontal transmission bias $T_A-T_B$, the coefficient $\beta_3$, and ratio $\beta_3/\beta_1$:
\begin{enumerate}
\item \emph{Fixation of the cooperative phenotype $A$}, if $T_A>T_B$ and $0<\beta_3$, or $T_A<T_B$ and $\beta_1<\beta_3$.

\item \emph{Fixation of the defector phenotype $B$}, if $T_A>T_B$ and $\beta_1<\beta_3<0$, or $T_A<T_B$ and $\beta_3<0$.

\item \emph{Protected polymorphism, or co-existence of both phenotypes}, if $T_A < T_B$ and $0<\beta_3<\beta_1$.

\item \emph{Fixation of either phenotype depending on initial frequency}, if $T_A>T_B$ and $\beta_3<\beta_1$.

\end{enumerate}

%\begin{figure}[H]
%  \centering
%  \begin{subfigure}{8cm}
%    \includegraphics[scale=0.5]{figure2a.pdf}
%    \caption{$v=1$, $T_A=T_B=T$, $\alpha \neq 0$}
%    \label{fig:results_a}
%  \end{subfigure}
%  \begin{subfigure}{8cm}
%    \includegraphics[scale=0.5]{figure2b.pdf}
%    \caption{$v=1$, $\alpha = 0$}
%    \label{fig:results_b}
%  \end{subfigure}
%  \begin{subfigure}{8cm}
%    \includegraphics[scale=0.5]{figure2c.pdf}
%    \caption{$v=0$}
%    \label{fig:results_c}
%  \end{subfigure}
%  \label{fig:results}
%  \caption{
%  \textbf{Numerical results for cultural evolution of cooperation.}
%  Shown are dynamics of \textbf{(a-b)} $\tilde{p}$, the frequency of parents with cooperative phenotype $A$; \textbf{(c)} $p'$, the frequency of adults with cooperative phenotype $A$.
%  The figure demonstrates fixation of cooperation (green), extinction of cooperation (blue)m and stable co-existence of cooperators and defectors (orange). %TODO ic (c), the blue and green got mixed. 
%  % TODO have all panels in one row.
%  }
%\end{figure}
